Information in recombination junctions can both compress and improve inferred tree sequences

### 2023-11-15 paper/writeup.tex
figure!
a few more notes

\bibliographystyle{plainnat}

@@

\section{Motivation and statement of problem}

@@

\section{Algorithm}

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction?

\subsection*{Notation}
    We begin with some notation. We define a \textit{tree sequence}, denoted $TS$, 
    as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
     $\left\{T_k\right\}_{k=1}^n$ is the set of trees,
     $N$ and $E$ are the sets of nodes and edges respectively,
     and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome.
     The length of each tree $T_k$ is some subinterval $[a_k,a_{k+1})$ 
     with the lengths of consecutive trees in the sequence 
     only intersecting each intervals endpoints. 
    Additionally, a node $x\in N$ is \textit{unary} if it only has one child, 
    ie. only a single edge extending below it on the tree.
    With these tree sequences our goal is to
    minimize the number of edges
    as well as infer coalescence of genes. 
    To remove edges we consider the following example.
    Suppose for a tree $T_k$ in the tree sequence $TS$ 
    there exists a branch which contains a unary node $b\in N$.
    Then the node $b$ has one parent, $a\in N$, 
    and one child $c\in N$,
    each with edges $e_{ab}, e_{bc}\in E$ connecting them.
    If there is an edge between the nodes $a$ and $c$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $e_{ab},e_{bc}$ 
    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}$. 
    This action reduces the length of the edge $e_{ac}$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such unary nodes 
    over the entire tree sequence.
\begin{center}
\begin{tikzpicture} 
	\node[circle,draw,inner sep=1pt] (a) at (0,2) {a}
	[growth parent anchor=south] 
		child { node[circle,draw,inner sep=1pt] {b} 		
	 			child {
	 				node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
	 					child{node (d) at (0,1) {$T_k$}
	 						edge from parent [draw=none]}
	 				edge from parent 
	 					node[left] {$e_{b,c}$}
	 			}	 		
	 		edge from parent
	 			node[left] {$e_{a,b}$}
	 	};

@@

\end{tikzpicture}
\end{center}
    In more detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$ 
    be a tree sequence.
    Within tskit we can track a tree sequence using a function called `EdgeDiffs'.
    This function tracks edges which are removed and added 
    between consecutive trees in the sequence. 
    As in \ref{} (tikzfigure), we can track if 
    $e_{a,b}$ and$e_{b,c}$ were removed from $T_k$ 
    when creating $T_{k+1}$, 
    and we can see if $e_{a,c}$ was just added in $T_{k+1}$.

@@

\section{Results (how it works)}

% (AVA)
% * reduction in edges ~ sequence length
% * speed increase ~ sequence length
% * proprotion of added edges that are true ~ sequeence length
% * apply to real data:
%     - % fewer edges
%     - % fewer distinct ancestors (ancestral paths?) of sample at some time

\section{Discrepancy Function}
% * How to measure agreement that includes haplotypes
% * Definition and Algorithm
% * Supp fig: runtime ~ # trees, samples
% * fig: how add edges rueduces discrepancy
%       - compare to Relate and tsinfer
% * Show: how add edges reduces discrepancy
% * THINK of a way to asynchronously computing discrepancy

@@

@@

\section{Data}
% * 1. Simulate data
% * 2. Run tsinfer
% * 3. Compare d(x,truth) for x in tsinfer, tsinfer+add edge, tsinfer+add edge + simplify, tsinfer+simplify

@@

@@

>>>>

@@

\subsection{Motivation and statement of problem}

@@

\begin{figure}
    \caption{
        Conceptual figure of what we're talking about.
        \label{fig:conceptual}
    }
\end{figure}

\section{Methods}

\subsection{Algorithm}

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)

\subsection*{Notation}
    We begin with some notation. We define a \textit{tree sequence}, denoted $TS$, 
    as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
     $\left\{T_k\right\}_{k=1}^n$ is the set of trees,
     $N$ and $E$ are the sets of nodes and edges respectively,
     and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome.
     The length of each tree $T_k$ is some subinterval $[a_k,a_{k+1})$ 
     with the lengths of consecutive trees in the sequence 
     only intersecting each intervals endpoints. 
    Additionally, a node $x\in N$ is \textit{unary} if it only has one child, 
    ie. only a single edge extending below it on the tree.
    With these tree sequences our goal is to
    minimize the number of edges
    as well as infer coalescence of genes. 
    To remove edges we consider the following example.
    Suppose for a tree $T_k$ in the tree sequence $TS$ 
    there exists a branch which contains a unary node $b\in N$.
    Then the node $b$ has one parent, $a\in N$, 
    and one child $c\in N$,
    each with edges $e_{ab}, e_{bc}\in E$ connecting them.
    If there is an edge between the nodes $a$ and $c$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $e_{ab},e_{bc}$ 
    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}$. 
    This action reduces the length of the edge $e_{ac}$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such unary nodes 
    over the entire tree sequence.

\begin{figure}
\begin{center}
\begin{tikzpicture} 
	\node[circle,draw,inner sep=1pt] (a) at (0,2) {a}
	[growth parent anchor=south] 
		child { node[circle,draw,inner sep=1pt] {b} 		
	 			child {
	 				node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
	 					child{node (d) at (0,1) {$T_k$}
	 						edge from parent [draw=none]}
	 				edge from parent 
	 					node[left] {$e_{b,c}$}
	 			}	 		
	 		edge from parent
	 			node[left] {$e_{a,b}$}
	 	};

@@

\end{tikzpicture}
\end{center}
\caption{
    \label{fig:extending_diagram}
}
\end{figure}`

    In more detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$ 
    be a tree sequence.
    Within tskit we can track a tree sequence using a function called `EdgeDiffs'.
    This function tracks edges which are removed and added 
    between consecutive trees in the sequence. 
    As in \ref{} (tikzfigure), we can track if 
    $e_{a,b}$ and$e_{b,c}$ were removed from $T_k$ 
    when creating $T_{k+1}$, 
    and we can see if $e_{a,c}$ was just added in $T_{k+1}$.

@@

\subsection{Discrepancy Function}
% * How to measure agreement that includes haplotypes
% * Definition and Algorithm
% * Supp fig: runtime ~ # trees, samples
% * fig: how add edges rueduces discrepancy
%       - compare to Relate and tsinfer
% * Show: how add edges reduces discrepancy
% * THINK of a way to asynchronously computing discrepancy

@@

\begin{figure}
    \caption{
        Conceptual figure displaying what discrepancy is?
        TODO: brainstorm.
        \label{fig:conceptual_discrepancy}
    }
\end{figure}

@@

\section{Results}

% (AVA)
% * reduction in edges ~ sequence length
% * speed increase ~ sequence length
% * proprotion of added edges that are true ~ sequeence length

\begin{figure}
    \caption{
        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
        as a function of number of trees (modulated by sequence length),
        on reasonably big simulated sequences.
        For "speed": plot ratio of runtime before/after extending.
        \label{fig:speed_and_edges}
    }
\end{figure}

Simple experiment:
(1) simulate, simplify, and extend;
(2) let $x$ be total matching span of simplified tree (here equal to the total span); $y$ total matching span of extended tree in truth and $z$ total not-matching span;
then $z$ is "total wrongly added stuff" and $y-x$ is "total rightly added stuff".

With tsinfer: (1) simulate and tsinfer; (2) do the above either with or without simplifying the tsinfer'ed trees.
(but also take into account that there's wrong stuff in tsinfer'ed tree: look at whether it reduces discrepancy!)

\begin{figure}
    \caption{
        Maybe: summaries of lengths and/or numbers of correctly and incorrectly added edges?
        (See notebook.)
        \label{fig:results_edges}
    }
\end{figure}

\begin{figure}
    \caption{
        Summaries of correctly and incorrectly added spans
        ... as a function of number of samples?
        \label{fig:results_discrepancy}
    }
\end{figure}

\subsection{Data}
% * 1. Simulate data
% * 2. Run tsinfer
% * 3. Compare d(x,truth) for x in tsinfer, tsinfer+add edge, tsinfer+add edge + simplify, tsinfer+simplify

@@

% * apply to real data:
%     - % fewer edges
%     - % fewer distinct ancestors (ancestral paths?) of sample at some time

\section{Discussion}

TODO

@@

\appendix

\begin{figure}
    \caption{
        Supplementary figure: runtime of discrepancy function as a function of number of trees
        (which we modulate by changing sequence length).
        \label{fig:speed_discrepancy}
    }
\end{figure}
